SimBio: a flexible and extensible Python package for systems modelling

### 2023-11-10 article.qmd
Rewrite of the introduction

Several tools already exist to define and simulate these systems.
In particular,
Biosimulators.org,
a registry of simulation tools,
list at least 15 Python softwares
among which are included
COPASI [@hoopsCOPASICOmplexPAthway2006],
Tellurium [@choiTelluriumExtensiblePythonbased2018],
GillesPy2 [@matthewGillesPy2BiochemicalModeling2023]
PySB [@lopezProgrammingBiologicalModels2013].
However,
some are only Python wrappers to external (C, C++) libraries.
They allow to "control" the model definition and simulation in a Python script,
but it is not possible to use existing Python packages
to change the way the simulation is carried out.
In other words, they are not extensible from Python.

avoid exporting to csv and loading in Python
large community

To write these models,
there are three main approaches.
Tools such as COPASI provide a Graphical User Interface (GUI),
where the components, relations and their values are defined.
Others use a Domain Specific Language (DSL),
such as Antimony,
which is a less verbose version of the System Biology Markup Language (SBML).
The last approach is to use a programming language such as Python,
which is the approach used by PySB.

Benefits of IDE integration (syntax highlighting, etc)

Here we present SimBio, a Python package to define and simulate reaction networks.
From the end-user point of view,
it is focused on providing an ergonomic experience
by integrating well with IDEs and static analysis tools
through the use of standard modern Python syntax.
From the power-user point of view,
it can be extended from Python,
since its coded in pure Python.
Finally,
from the developer/maintainer point of view,
it has a modular architecture with a clear separation of concerns,
making it easier to maintain or to contribute new code.

>>>>

Several tools already exist to define and simulate these systems.
In particular,
BioSimulators.org [@shaikhBioSimulatorsCentralRegistry2022],
a registry of simulation tools,
list at least 15 Python softwares.
Some of them are:

1. COPASI [@hoopsCOPASICOmplexPAthway2006],
a stand-alone software
which provides a Graphical User Interface (GUI) to define and simulate models.
2. Tellurium [@choiTelluriumExtensiblePythonbased2018],
which defines itself as a Python-based modeling environment
and uses a C++ library called libRoadRunner [@welshLibRoadRunnerHighPerformance2023] to simulate them.
3. PySB [@lopezProgrammingBiologicalModels2013],
a Python library
that created a DSL using standard Python to define models
which are then compiled to ODE using a Perl library called BioNetGen [@harrisBioNetGenAdvancesRulebased2016].

A limitation of these tools is their extensibility from Python,
as they are or use wrapped libraries in other languages.
As such, they allow to "control" the model definition and simulation in a Python script,
but we cannot leverage other packages from the huge Python community
to change the way the simulation is carried out.
In some sense, we could say that
they facilitate not having to export the simulation result to a CSV file
and then loading it in Python to continue the analysis.

Another weak point is related to the way the models are written.
While most tools support standards such as the System Biology Markup Language (SBML) [@huckaSBMLL3V2],
no one is expected to write an SBML file directly.
Tellurium uses a Domain Specific Language (DSL) called Antimony [@smithAntimonyModularModel2009],
which can be translated to SBML.
A DSL permits to reuse of the code *as is* in different programming environments,
but will not be recognized in Integrated Development Enviromnents (IDE)
and, then, cannot provide the development experience we have come to expect:
syntax highlighting, code completion, refactoring, static analysis, among others.
This has been recognized by the community,
and an extension for Visual Studio Code has been created as part of a master thesis [@vscode-antimony].
But maintaining it into the future could be burdensome for the smaller systems biology community.

To address these two issues,
we developed SimBio,
a Python package to define and simulate reaction networks.
From the end-user point of view,
it is focused on providing an ergonomic experience
by integrating well with IDEs and static analysis tools
through the use of standard modern Python syntax.
From the power-user point of view,
it can be extended from Python,
since its coded in pure Python.
Finally,
from the developer/maintainer point of view,
it has a modular architecture with a clear separation of concerns,
making it easier to maintain or to contribute new code.
